Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-10-07 paper/oopsla.tex
added example showing the need for resptec
AEC draft

Recently, \citet{promising-ldrf} showed that certain combinations of compiler
optimizations are inconsistent with local \drf{} guarantees.  All of the
examples that prove inconsistency have the same shape: they combine read
introduction and case analysis (aka, if-closure).  Effectively, this turns
one read into two, where different conditional branches can be taken for the
two copies of the read.  This is reminiscent of the type of \emph{bait and
  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
\begin{gather*}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI\IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    % \\[-1ex]
    % \hbox{\begin{tikzinline}[node distance=1.5em]
    %     \event{rx}{\DR{x}{1}}{}
    %     \event{wy}{\DW{y}{1}}{right=of rx}
    %     \po{rx}{wy}
    %     \event{ry}{\DR{y}{1}}{right=3em of wy} 
    %     \event{wx}{\DW{x}{1}}{right=of ry}
    %     \event{wz}{\DW{z}{1}}{right=of wx}
    %     \event{rb}{\DR{b}{1}}{right=of wz}
    %     \event{wb1}{\DW{b}{1}}{right=3em of rb}
    %     \po{ry}{wx}
    %     \rf{wb1}{rb}
    %     \rf{wy}{ry}
    %     \rf[out=-170,in=-10]{wx}{rx}
    %     \po{rb}{wz}
    %     \po[out=15,in=165]{ry}{wz}
    %   \end{tikzinline}}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{rb}{\DR{b}{1}}{right=3em of wy}
        \event{ry}{\DR{y}{1}}{right=of rb} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{wb1}{\DW{b}{1}}{right=3em of wz}
        \po{ry}{wx}
        \rf[out=-170,in=-10]{wb1}{rb}
        \rf[out=15,in=165]{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po[out=15,in=165]{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
Under all variants of \PwT{}, this outcome is disallowed, due to the cycle
involving $x$ and $y$.\footnote{All of the reads in \ref{oota4} are
  cross-thread, so there is no difference between \PwTmca{1} and \PwTmca{2}.
  For \PwTc, there is a cycle in ${\rrfx}\cup{\lt}$.}  Under \PS{}, this
outcome is allowed by {baiting} with the \texttt{else} branch, then
{switching} to the \texttt{then} branch, based on a coin flip (${b}$).
% Using a combination of read introduction and case analysis,
% the middle thread can be rewritten to:
% \begin{displaymath}
%   \begin{array}{l}
%     \PR{y}{r} \SEMI \IF{b}\THEN\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \ELSE\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \FI
%   \end{array}
% \end{displaymath}
% \footnote{Call the threads \texttt{s}, \texttt{t}, and
%   \texttt{u}.  To get the result in the promising semantics, first execute
%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
%   removing timestamps and replacing the word \emph{promise} by
%   \emph{speculation}.}
% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
% of temporal reasoning could result in violations of the Java security
% architecture.

@@

\citeauthor{promising-ldrf} argue that the outcome of \ref{LDRF-Fail-PS} is
inevitable due to compiler optimizations.  The examples crucially involve the
following sequence of operations:
\begin{itemize}
\item read introduction,
\item if introduction, branching on the read just introduced.
\end{itemize}
We believe this combination of optimizations is unsound.  This is obviously
the case in \cXI: read introduction may cause undefined behavior (\ub), due to
the possible introduction of a data race.

The situation is more delicate in \llvm.  The short version of the story is
that load-hoisting followed by case analysis is unsound in \llvm, without freeze. 
This happens because:
\begin{itemize}
\item read introduction may result in the undefined value $\UNDEFINED$, due
  to the possible introduction of a data race \cite{DBLP:conf/cgo/ChakrabortyV17}, and
\item branching on an undefined value in \llvm{} results in \ub. 
\end{itemize}
\llvm{} delays \ub{} using the undefined value.  This allows \llvm{} to
perform optimizations such as load hoisting, where
\begin{math}
  \IF{C}\THEN \PR{x}{r}\FI
\end{math}
is rewritten to 
\begin{math}
  \PR{x}{s}\SEMI
  \LET{r}{\TERNARY{C}{s}{r}}.
\end{math}
Despite this, other optimizations regularly performed by \llvm{} are
unsound \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  An example is loop switching,
where
\begin{math}
  \WHILE{C_1}\DO \IF{C_2}\THEN \aCmd_1 \ELSE \aCmd_2 \FI \OD
\end{math}
is rewritten to 
\begin{math}
  \IF{C_2}\THEN \WHILE{C_1}\DO \aCmd_1 \OD \ELSE \WHILE{C_1}\DO \aCmd_2 \FI \OD.
\end{math}
Freeze was introduced in \llvm{} in order to make such optimizations sound by
allowing branch on frozen $\UNDEFINED$ to give nondeterministic choice rather
than \ub{}.  In the RFC for freeze, \citet{nuno} says: \emph{``Note that
  having branch on poison not trigger \ub{} has its own problems.  We believe
  this is a good tradeoff.''}  \ref{LDRF-Fail-PS} demonstrates a concrete
problem with this tradeoff.  Other compilers, such as Compcert, are more
conservative \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.

Thus, the difference between \PS{} and \PwT{} can be understood in terms of
the valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis on the value read.  \PwT{} validates case analysis,
but invalidates read introduction.

@@

\begin{scope}
  The substitution $[x/r]$ does not work with Skolemization, even for the
  dependent case, since we lose the unique marker for each read.  In effect,
  this forces all reads of a location to see the same values.
  % To be concrete, the candidate
  % definition would modify \ref{L4} to be:
  % \begin{enumerate}
  % \item[\ref{L4})]
  %   $\aTr{\bEvs}{\bForm} \riff \aVal{=}\aLoc\limplies\bForm[\aLoc/\aReg]$.
  %   % \item[\ref{L5})]
  %   %   $\aTr{\cEvs}{\bForm} \riff (\aVal{=}\aLoc\lor\TRUE)\limplies\bForm[\aLoc/\aReg]$. %, when $\aEvs\neq\emptyset$,
  %   % \item[\ref{L6})] 
  %   %   $\aTr{\dEvs}{\bForm}\; \riff \bForm$, when $\aEvs=\emptyset$.
  % \end{enumerate}
  Using this definition, consider the following:
  \begin{gather*}
    \PR{x}{r}\SEMI
    \PR{x}{s}\SEMI
    \IF{r{<}s}\THEN \PW{y}{1}\FI 
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
        \event{a1}{\DR{x}{1}}{}
        \event{a2}{\DR{x}{2}}{right=of a1}
        \event{a3}{1{=}x\limplies 2{=}x\limplies x{<} x\mid\DW{y}{1}}{right=of a2}
        \po[out=20,in=160]{a1}{a3}
        \po{a2}{a3}
      \end{tikzinline}}
  \end{gather*}
  Although the execution seems reasonable, the precondition on the write is
  not a tautology.
\end{scope}

@@

\jjr{} enforces downset closure in the prefixing rule.  Even without this,
downset closure would be different for the two semantics, due to the use of
substitution in \jjr{}.  Consider the final pomset in the last example of
\textsection\ref{sec:downset} under the semantics of this paper, which elides
the middle read event:
\begin{align*}
  \begin{gathered}[t]
    \PW{x}{0} 
    \SEMI\PR{x}{r} 
    \SEMI\IF{r{\geq}0}\THEN \PW{y}{1} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a0}{\DW{x}{0}}{}
        % \event{a1}{\DR{x}{1}}{right=of a0}
        \event{a2}{r{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
        % \wk{a0}{a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
In \jjr{}, the substitution $[x/r]$ is performed by the middle read
regardless of whether it is included in the pomset, with the subsequent
substitution of $[0/x]$ by the preceding write, we have $[x/r][0/x]$, which
is $[0/r][0/x]$, resulting in:
\begin{align*}
  \begin{gathered}[t]
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a0}{\DW{x}{0}}{}
        % \event{a1}{\DR{x}{1}}{right=of a0}
        \event{a2}{0{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
        % \wk{a0}{a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}

\myparagraph{Consistency}
\jjr{} imposes \emph{consistency}, which requires that for every pomset
$\aPS$, $\bigwedge_{\aEv}\labelingForm(\aEv)$ is satisfiable.  
\begin{scope}
  Associativity requires that we allow pomsets with inconsistent
  preconditions.  Consider a variant of the example from \textsection\ref{sec:semca}.
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\mid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\mid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\mid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}
  Associating left and right, we have:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}  
  Associating into the middle, instead, we require:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\mid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
            \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\mid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}
  Joining left and right, we have:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{x}{1}}{}
            \event{b}{\DW{y}{1}}{right=of a}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}  
\end{scope}

>>>>

Recently, \citet{promising-ldrf} showed that certain combinations of compiler
optimizations are inconsistent with local \drf{} guarantees.  All of the
examples that prove inconsistency have the same shape: they combine
read-introduction and case analysis (aka, if-introduction).  Effectively,
this turns one read into two, where different conditional branches can be
taken for the two copies of the read.  This is reminiscent of the type of
\emph{bait and switch} behavior noted by
\citet{DBLP:journals/pacmpl/JagadeesanJR20}: the promising semantics (\PS{})
\cite{DBLP:conf/popl/KangHLVD17} and related models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
\begin{gather*}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI\IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    % \\[-1ex]
    % \hbox{\begin{tikzinline}[node distance=1.5em]
    %     \event{rx}{\DR{x}{1}}{}
    %     \event{wy}{\DW{y}{1}}{right=of rx}
    %     \po{rx}{wy}
    %     \event{ry}{\DR{y}{1}}{right=3em of wy} 
    %     \event{wx}{\DW{x}{1}}{right=of ry}
    %     \event{wz}{\DW{z}{1}}{right=of wx}
    %     \event{rb}{\DR{b}{1}}{right=of wz}
    %     \event{wb1}{\DW{b}{1}}{right=3em of rb}
    %     \po{ry}{wx}
    %     \rf{wb1}{rb}
    %     \rf{wy}{ry}
    %     \rf[out=-170,in=-10]{wx}{rx}
    %     \po{rb}{wz}
    %     \po[out=15,in=165]{ry}{wz}
    %   \end{tikzinline}}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{rb}{\DR{b}{1}}{right=3em of wy}
        \event{ry}{\DR{y}{1}}{right=of rb} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{wb1}{\DW{b}{1}}{right=3em of wz}
        \po{ry}{wx}
        \rf[out=-170,in=-10]{wb1}{rb}
        \rf[out=15,in=165]{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po[out=15,in=165]{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
Under all variants of \PwT{}, this outcome is disallowed, due to the cycle
involving $x$ and $y$.\footnote{All of the reads in \ref{oota4} are
  cross-thread, so there is no difference between \PwTmca{1} and \PwTmca{2}.
  For \PwTc, there is a cycle in ${\rrfx}\cup{\lt}$.}  Under \PS{}, this
outcome is allowed by {baiting} with the \texttt{else} branch, then
{switching} to the \texttt{then} branch, based on a coin flip (${b}$).
% Using a combination of read introduction and case analysis,
% the middle thread can be rewritten to:
% \begin{displaymath}
%   \begin{array}{l}
%     \PR{y}{r} \SEMI \IF{b}\THEN\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \ELSE\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \FI
%   \end{array}
% \end{displaymath}
% \footnote{Call the threads \texttt{s}, \texttt{t}, and
%   \texttt{u}.  To get the result in the promising semantics, first execute
%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
%   removing timestamps and replacing the word \emph{promise} by
%   \emph{speculation}.}
% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
% of temporal reasoning could result in violations of the Java security
% architecture.

@@

\citeauthor{promising-ldrf} argue that the outcome of \ref{LDRF-Fail-PS} is
inevitable due to compiler optimizations.  The examples crucially involve the
following sequence of operations:
\begin{itemize}
\item read-introduction,
\item if-introduction, branching on the read just introduced.
\end{itemize}
We believe this combination of optimizations is unsound.  This is obviously
the case in \cXI: read-introduction may cause undefined behavior (\ub), due to
the possible introduction of a data race.

The situation is more delicate in \llvm.  The short version of the story is
that load-hoisting followed by case analysis is unsound in \llvm, without freeze. 
This happens because:
\begin{itemize}
\item read-introduction may result in the undefined value $\UNDEFINED$, due
  to the possible introduction of a data race \cite{DBLP:conf/cgo/ChakrabortyV17}, and
\item branching on an undefined value in \llvm{} results in \ub. 
\end{itemize}
\llvm{} delays \ub{} using the undefined value.  This allows \llvm{} to
perform optimizations such as load hoisting, where
\begin{math}
  \IF{C}\THEN \PR{x}{r}\FI
\end{math}
is rewritten to 
\begin{math}
  \PR{x}{s}\SEMI
  \LET{r}{\TERNARY{C}{s}{r}}.
\end{math}
Despite this, other optimizations regularly performed by \llvm{} are
unsound \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  An example is loop switching,
where
\begin{math}
  \WHILE{C_1}\DO \IF{C_2}\THEN \aCmd_1 \ELSE \aCmd_2 \FI \OD
\end{math}
is rewritten to 
\begin{math}
  \IF{C_2}\THEN \WHILE{C_1}\DO \aCmd_1 \OD \ELSE \WHILE{C_1}\DO \aCmd_2 \FI \OD.
\end{math}
Freeze was introduced in \llvm{} in order to make such optimizations sound by
allowing branch on frozen $\UNDEFINED$ to give nondeterministic choice rather
than \ub{}.  In the RFC for freeze, \citet{nuno} says: \emph{``Note that
  having branch on poison not trigger \ub{} has its own problems.  We believe
  this is a good tradeoff.''}  \ref{LDRF-Fail-PS} demonstrates a concrete
problem with this tradeoff.  Other compilers, such as Compcert, are more
conservative \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.

Thus, the difference between \PS{} and \PwT{} can be understood in terms of
the valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis on the value read.  \PwT{} validates case analysis,
but invalidates read-introduction.

@@

\begin{scope}
  The substitution $[x/r]$ does not work with Skolemization, even for the
  dependent case, since we lose the unique marker for each read.  In effect,
  this forces all reads of a location to see the same values.
  % To be concrete, the candidate
  % definition would modify \ref{L4} to be:
  % \begin{enumerate}
  % \item[\ref{L4})]
  %   $\aTr{\bEvs}{\bForm} \riff \aVal{=}\aLoc\limplies\bForm[\aLoc/\aReg]$.
  %   % \item[\ref{L5})]
  %   %   $\aTr{\cEvs}{\bForm} \riff (\aVal{=}\aLoc\lor\TRUE)\limplies\bForm[\aLoc/\aReg]$. %, when $\aEvs\neq\emptyset$,
  %   % \item[\ref{L6})] 
  %   %   $\aTr{\dEvs}{\bForm}\; \riff \bForm$, when $\aEvs=\emptyset$.
  % \end{enumerate}
  Using this definition, consider the following:
  \begin{gather*}
    \PR{x}{r}\SEMI
    \PR{x}{s}\SEMI
    \IF{r{<}s}\THEN \PW{y}{1}\FI 
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
        \event{a1}{\DR{x}{1}}{}
        \event{a2}{\DR{x}{2}}{right=of a1}
        \event{a3}{1{=}x\limplies 2{=}x\limplies x{<} x\bigmid\DW{y}{1}}{right=of a2}
        \po[out=20,in=160]{a1}{a3}
        \po{a2}{a3}
      \end{tikzinline}}
  \end{gather*}
  Although the execution seems reasonable, the precondition on the write is
  not a tautology.
\end{scope}

@@

\jjr{} enforces downset closure in the prefixing rule.  Even without this,
downset closure would be different for the two semantics, due to the use of
substitution in \jjr{}.  Consider the final pomset in the last example of
\textsection\ref{sec:downset} under the semantics of this paper, which elides
the middle read event:
\begin{align*}
  \begin{gathered}[t]
    \PW{x}{0} 
    \SEMI\PR{x}{r} 
    \SEMI\IF{r{\geq}0}\THEN \PW{y}{1} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a0}{\DW{x}{0}}{}
        % \event{a1}{\DR{x}{1}}{right=of a0}
        \event{a2}{r{\geq}0\bigmid\DW{y}{1}}{right=3em of a1}      
        % \wk{a0}{a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
In \jjr{}, the substitution $[x/r]$ is performed by the middle read
regardless of whether it is included in the pomset, with the subsequent
substitution of $[0/x]$ by the preceding write, we have $[x/r][0/x]$, which
is $[0/r][0/x]$, resulting in:
\begin{align*}
  \begin{gathered}[t]
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a0}{\DW{x}{0}}{}
        % \event{a1}{\DR{x}{1}}{right=of a0}
        \event{a2}{0{\geq}0\bigmid\DW{y}{1}}{right=3em of a1}      
        % \wk{a0}{a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}

\myparagraph{Consistency}
\jjr{} imposes \emph{consistency}, which requires that for every pomset
$\aPS$, $\bigwedge_{\aEv}\labelingForm(\aEv)$ is satisfiable.  
\begin{scope}
  Associativity requires that we allow pomsets with inconsistent
  preconditions.  Consider a variant of the example from \textsection\ref{sec:semca}.
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\bigmid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\bigmid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\bigmid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\bigmid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}
  Associating left and right, we have:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}  
  Associating into the middle, instead, we require:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\aExp\bigmid\DW{x}{1}}{}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\bigmid\DW{x}{1}}{}
            \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
          \end{tikzinline}}
      \end{gathered}
      &&
      \begin{gathered}
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\lnot\aExp\bigmid\DW{y}{1}}{}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}
  Joining left and right, we have:
  \begin{scope}
    \footnotesize
    \begin{align*}
      \begin{gathered}
        \IF{\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
        \SEMI
        \IF{\aExp}\THEN\PW{y}{1}\FI
        \SEMI
        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
        \\
        \hbox{\begin{tikzinline}[node distance=1em]
            \event{a}{\DW{x}{1}}{}
            \event{b}{\DW{y}{1}}{right=of a}
          \end{tikzinline}}
      \end{gathered}
    \end{align*}
  \end{scope}  
\end{scope}

\myparagraph{Causal Strengthening}
% \labeltext[]{Causal Strengthening}{xCausal}
\jjr{} imposes \emph{causal strengthening}, which requires for every pomset
$\aPS$, if $\bEv\lt\aEv$ then $\labelingForm(\aEv) \rimplies \labelingForm(\bEv)$. 
\begin{scope}
  Associativity requires that we allow pomsets without causal strengthening.
  Consider the following.
  \begin{align*}
    \begin{gathered}
      \IF{\aExp}\THEN\PR{x}{r}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\aExp\bigmid\DR{x}{1}}{}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \PW{y}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{r{=}1\bigmid\DW{y}{1}}{}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\lnot\aExp\bigmid\DR{x}{1}}{}
        \end{tikzinline}}
    \end{gathered}
  \end{align*}
  Associating left, with causal strengthening:
  \begin{align*}
    \begin{gathered}
      \IF{\aExp}\THEN\PR{x}{r}\FI
      \SEMI
      \PW{y}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\aExp\bigmid\DR{x}{1}}{}
          \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
          \po{a}{b}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\lnot\aExp\bigmid\DR{x}{1}}{}
        \end{tikzinline}}
    \end{gathered}
  \end{align*}
  Finally, merging:
  \begin{align*}
    \begin{gathered}
      \IF{\aExp}\THEN\PR{x}{r}\FI
      \SEMI
      \PW{y}{r}
      \SEMI
      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\DR{x}{1}}{}
          \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
          \po{a}{b}
        \end{tikzinline}}
    \end{gathered}
  \end{align*}
  Instead, associating right:
  \begin{align*}
    \begin{gathered}
      \IF{\aExp}\THEN\PR{x}{r}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\aExp\bigmid\DR{x}{1}}{}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \PW{y}{r}
      \SEMI
      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\lnot\aExp\bigmid\DR{x}{1}}{}
          \event{b}{r{=}1\bigmid\DW{y}{1}}{left=of a}
        \end{tikzinline}}
    \end{gathered}
  \end{align*}
  Merging:
  \begin{align*}
    \begin{gathered}
      \IF{\aExp}\THEN\PR{x}{r}\FI
      \SEMI
      \PW{y}{r}
      \SEMI
      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
      \\
      \hbox{\begin{tikzinline}[node distance=1em]
          \event{a}{\DR{x}{1}}{}
          \event{b}{\DW{y}{1}}{right=of a}
          \po{a}{b}
        \end{tikzinline}}
    \end{gathered}
  \end{align*}
  With causal strengthening, the precondition of $\DW{y}{1}$ depends upon how
  we associate.  This is not an issue in \jjr{}, which always associates to
  the right.
\end{scope}

### 2021-10-07 paper/outro.tex
AEC updates
AEC draft

\PwTmca{} does not validate access elimination: store-forwarding and
dead-write-removal are unsound.  We expect that these can be validated by
allowing events with different labels to merge. % in some instances.

@@

% In a similar vein,
\PwT{} does not validate read introduction, whereas speculative operational
semantics do.  Recent work shows a tension between read introduction and
compositional reasoning for temporal safety properties (see
\textsection\ref{sec:promising}).  Nonetheless, read introduction is
ubiquitous is some compilers.  It would be interesting to know if there is a
performance penalty for banning read introduction.

>>>>

\PwTmca{} does not validate access elimination: store-forwarding and
dead-write-removal are unsound.  We expect that these can be validated by
allowing events with different actions to merge. % in some instances.

@@

% In a similar vein,
\PwT{} does not validate read introduction, whereas speculative operational
semantics do.  Recent work shows a tension between read introduction and
compositional reasoning for temporal safety properties (see
\textsection\ref{sec:promising}).  Nonetheless, read introduction is
ubiquitous in some compilers.  It would be interesting to know if there is a
performance penalty for banning read introduction.
